Clustering of Last.fm Music

### 2015-05-07 writeup/MSDgroupwriteup.tex
small edits

For our project, we explored various song recommendation systems algorithms such as popularity recommendations, artist based recommensations, matrix factorization, and user and item based similarity measures.  
% memory based collaborative filtering

@@

We choose to subset the data to users who have listened to more than 27 songs and songs which were listened to by more than 22 users. This resulted in a datset which was 10\% the size of the original. Despite subsetting the data, we are still dealing with a matrix of $8130$ users by $11861$ songs, so some of our algorithms will take a fair amount of time to run.

@@

Specifically, we used MAP@500 scores which take into account the first 500 recommendations given to each user. Generally, the score looks at the precision between a list of recommendations for a user and that user's test set. This was done by calculating the average precision at each element in the list of recommended items (i.e. percentage of correct items among first k recommendations) and averaging these for the first 500 recommendations (i.e. for each $k \epsilon {1, ..., 500}$, calculate the precision and average them together).
We then averaged the score over all users. Intuitively, the score looks at the percentage of recommendations that were in-line with the test set of songs actually listened to but also takes order into account. So it is preferable to recommend the songs in the user's test set earlier in the list of 500. The MAP score is similar to the AUC of the ROC curve discussed in class except it takes ordering of the recommendations into account, which is an important factor to consider in a recommendation system.

@@

\emph{Note}: $\lambda = 10$ was used and $\Omega$ is the set of all indices in the matrix which have an observation.
We keep track of the root mean square error (RMSE) versus the testing set (i.e. how close our prediction is as compared to the actual normalized play count in the testing set) and the log joint likelihood of the training set as a function of iteration (see Figures \ref{fig:rmseplot} and \ref{fig:likeplot}).
There are several hyper parameters that can be tweaked in order to achieve a better mAP result using matrix factorization. The main ones that  was varied were the variance, the rank d of the factorization, and the iterations. As the project was being done using Python and the main computational package (\emph{scikit-learn}) did not have probabilistic matrix factorization functions, we decided to apply the algorithm that we had wrote ourselves for a different class.

@@

In order to tackle the issue of using implicit user feedback, we tried four different scaling schemas independently. These include normalizing user play counts by the total number of plays for the user, normalizing song play counts by the total number of plays for the song, making the data binary (changing all non-zero play counts to 1), and also applying tfidf. While testing out these schemas, we continued to vary the hyperparameters d and variance in order to determine if they changed the MAP results significantly.

From the experiments, it was found that the rank d did not contribute significantly to the MAP values especially in settings where the MAP values were close to the baseline. Higher variance values were found to actually significantly increase the MAP values. In terms of scaling schemas, making the data binary produced the best result, with the second best being normalizing the data against the total number of plays for each user. Perhaps it can be said that using number of plays introduces more noise than simply considering if the user has played the song or not. Interestingly enough, normalizing the data against total number of plays for each song performed the worst and this most likely relates to the fact that number of plays is more meaningful relative to each user rather than to each song.

>>>>

For our project, we explored various song recommendation systems algorithms such as popularity recommendations, artist based recommendations, matrix factorization, and user and item based similarity measures.  
% memory based collaborative filtering

@@

We choose to subset the data to users who have listened to more than 27 songs and songs which were listened to by more than 22 users. This resulted in a dataset which was 10\% the size of the original. Despite subsetting the data, we are still dealing with a matrix of $8130$ users by $11861$ songs, so some of our algorithms will take a fair amount of time to run.

@@

Specifically, we used MAP@500 scores which take into account the first 500 recommendations given to each user. Generally, the score looks at the precision between a list of recommendations for a user and that user's test set. This was done by calculating the average precision at each element in the list of recommended items (i.e. percentage of correct items among first $k$ recommendations) and averaging these for the first 500 recommendations (i.e. for each $k \in \{1, ..., 500\}$, calculate the precision and average them together).
We then averaged the score over all users. Intuitively, the score looks at the percentage of recommendations that were in-line with the test set of songs actually listened to but also takes order into account. So it is preferable to recommend the songs in the user's test set earlier in the list of 500. The MAP score is similar to the AUC of the ROC curve discussed in class except it takes ordering of the recommendations into account, which is an important factor to consider in a recommendation system.

@@

\emph{Note: $\lambda = 10$ was used and $\Omega$ is the set of all indices in the matrix which have an observation.}
%We keep track of the root mean square error (RMSE) versus the testing set (i.e. how close our prediction is as compared to the actual normalized play count in the testing set) and the log joint likelihood of the training set as a function of iteration (see Figures \ref{fig:rmseplot} and \ref{fig:likeplot}).

There are several hyper parameters that can be tweaked in order to achieve a better MAP result using matrix factorization. The main ones that  was varied were the variance, the rank d of the factorization, and the iterations. As the project was being done using Python and the main computational package (\emph{scikit-learn}) did not have probabilistic matrix factorization functions, we decided to apply the algorithm that we had wrote ourselves for a different class.

@@

In order to tackle the issue of using implicit user feedback, we tried four different scaling schemas independently. These include normalizing user play counts by the total number of plays for the user, normalizing song play counts by the total number of plays for the song, making the data binary (changing all non-zero play counts to 1), and also applying tfidf. While testing out these schemas, we continued to vary the hyper-parameters: $d$ and $\sigma$, in order to determine if they changed the MAP results significantly.

From the experiments, it was found that the rank $d$ did not contribute significantly to the MAP values especially in settings where the MAP values were close to the baseline. Higher variance values were found to actually significantly increase the MAP values. In terms of scaling schemas, making the data binary produced the best result, with the second best being normalizing the data against the total number of plays for each user. Perhaps it can be said that using number of plays introduces more noise than simply considering if the user has played the song or not. Interestingly enough, normalizing the data against total number of plays for each song performed the worst and this most likely relates to the fact that number of plays is more meaningful relative to each user rather than to each song.
